---
title: "Do Higher Temperatures Causally Affect Walmart Weekly Sales?"
author: "Ashar Hashmi"
format:
  pdf:
    fontsize: 11pt
    geometry: margin=1in
    toc: true
    number-sections: true
    code-overflow: wrap
execute:
  echo: true
  warning: false
  message: false
---

1.  Introduction

Walmart wants to know whether rising temperatures might hurt or help their 
weekly sales. With temperatures going up over time, this question matters 
for planning and forecasting.

I have weekly data from 45 Walmart stores covering February 2010 through 
October 2012. For each store and week, I observe:

-   Weekly sales
-   Temperature
-   Whether it was a holiday week
-   Fuel price, CPI, and unemployment

Research Question: In these stores, do higher temperatures causally affect
weekly sales? The main challenge is that both temperature and sales 
are seasonal. For instance, temperatures are higher in summer, and sales also 
change across the year because of holidays and shopping patterns. 
If we simply compare hot weeks to cold weeks, we might mistake seasonality 
for a temperature effect.

To handle this, I use a panel data approach with:

- Store fixed effects: They control for any permanent differences between 
stores (like size, location, customer base).

- Date fixed effects: They control for anything that happens in the same week 
for all stores, such as national holidays, seasonality, or macroeconomic shocks.

- Two‑way clustered standard errors by Store and Date: This is standard in 
store‑week panel settings.

With store fixed effects and week fixed effects, the model is using changes 
within the same store over time, after removing anything that is common to all 
stores in the same week like national shopping patterns, holidays, 
or seasonal demand. So the temperature effect is identified from week-to-week 
temperature changes for a given store, not from differences between stores.

```{r}

# 2. Load libraries

library(tidyverse)
library(lubridate)
library(fixest)
library(modelsummary)
library(scales)
library(knitr)
library(kableExtra)

options(scipen = 999)
set.seed(123)

```

```{r}

# 3. Load the Data

walmart_raw <- readr::read_csv(
  "data/Walmart_Sales.csv",
  show_col_types = FALSE
)

# Show structure 
str(walmart_raw)

# Show first 6 rows 
head(walmart_raw)

# Summary statistics 
summary(walmart_raw)

```

```{r}

# 4. Data Cleaning and Preparation

# 4.1 Fixing the Date column
# The dataset contains dates written in different 
# formats (some use dashes, some slashes).
# To avoid mistakes, I parse dates using multiple possible formats.

walmart <- walmart_raw %>%
  mutate(
    Date = parse_date_time(Date, orders = c("dmy", "mdy", "dmY", "mdY"))
  ) %>%
  filter(!is.na(Date)) %>%
  mutate(Date = as_date(Date)) %>%
  arrange(Store, Date)

# Quick date checks
range(walmart$Date)

# Check that observations are weekly within store
walmart %>%
  group_by(Store) %>%
  arrange(Date) %>%                
  summarise(
    median_gap_days = median(as.numeric(Date - lag(Date)), na.rm = TRUE),
    pct_gap_7days = mean(as.numeric(Date - lag(Date)) == 7, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  summary()

```

The median gap is 7 days and most gaps are 7 days, so the dates 
are correctly spaced.

```{r}

# 4.2 Panel structure check

cat("Rows:", nrow(walmart), "\n")
cat("Stores:", n_distinct(walmart$Store), "\n")
cat("Weeks:", n_distinct(walmart$Date), "\n")

walmart %>%
  group_by(Store) %>%
  summarise(n_weeks = n_distinct(Date), .groups = "drop") %>%
  summary()

```

All stores appear in most weeks which shows a balanced panel.

```{r}

# 4.3 Log sales

walmart <- walmart %>%
  mutate(Log_Sales = log(Weekly_Sales))

```

Weekly sales are large and skewed. Taking logs makes the distribution more 
stable and turns coefficients into approximate percent changes, 
which is easier to interpret.

```{r}

# 4.4 Note about Holiday_Flag with Date fixed effects

walmart %>%
  group_by(Date) %>%
  summarise(n_holiday_values = n_distinct(Holiday_Flag), .groups = "drop") %>%
  count(n_holiday_values)

```

In this dataset, Holiday_Flag is the same for all stores in the same week.

If we include Date fixed effects, Holiday_Flag is perfectly collinear 
(absorbed by the date dummies). We can quickly verify that.

Because every date has only one holiday value, I do not include Holiday_Flag 
separately in models that have Date FE.

```{r}

# 5. Summary Statistics

walmart %>%
  summarise(
    mean_sales = mean(Weekly_Sales, na.rm = TRUE),
    sd_sales   = sd(Weekly_Sales, na.rm = TRUE),
    mean_temp  = mean(Temperature, na.rm = TRUE),
    sd_temp    = sd(Temperature, na.rm = TRUE),
    mean_fuel  = mean(Fuel_Price, na.rm = TRUE),
    mean_cpi   = mean(CPI, na.rm = TRUE),
    mean_unemp = mean(Unemployment, na.rm = TRUE)
  )

```


```{r}

# 6. Exploratory Graphs – Guiding the Model
# These graphs help us see why strong time controls are needed.

# 6.1 Temperature over time

ggplot(walmart, aes(x = Date, y = Temperature, group = Store)) +
  geom_line(alpha = 0.10, color = "steelblue") +
  geom_smooth(aes(group = 1), method = "loess", se = TRUE, color = "darkred") +
  labs(
    title = "Temperature Over Time (All Stores)",
    subtitle = "Temperature has a clear seasonal pattern.",
    x = "Week",
    y = "Temperature (°F)"
  ) +
  theme_minimal()

```

Temperature peaks in summer and lowers in winter, it shows a strong seasonal cycle.

```{r}

# 6.2 Sales over time

ggplot(walmart, aes(x = Date, y = Weekly_Sales, group = Store)) +
  geom_line(alpha = 0.10, color = "darkgreen") +
  geom_smooth(aes(group = 1), method = "loess", se = TRUE, color = "darkred") +
  labs(
    title = "Weekly Sales Over Time (All Stores)",
    subtitle = "Sales also vary across the year, with holiday spikes.",
    x = "Week",
    y = "Weekly Sales"
  ) +
  scale_y_continuous(labels = dollar) +
  theme_minimal()

```

Sales also have seasonal patterns and big holiday bumps.
If we want the pure temperature effect, we must control for these common 
time patterns, which is exactly what date fixed effects do.

```{r}

# 6.3 Raw relationship between sales and temperature

ggplot(walmart, aes(x = Temperature, y = Weekly_Sales)) +
  geom_point(alpha = 0.05, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  labs(
    title = "Raw Sales vs. Temperature",
    subtitle = "Without controls, sales look higher when it's warmer,
    but that could be seasonality.",
    x = "Temperature (°F)",
    y = "Weekly Sales"
  ) +
  scale_y_continuous(labels = dollar) +
  theme_minimal()

```

The raw plot suggests higher sales in warmer weeks.
But we already know both variables move with seasons so this is likely spurious.
The scatter also looks slightly curved, so I later test a quadratic
specification to see whether the relationship departs from linearity 
once fixed effects are included.

7. Thinking About Causal Identification
To answer “causally”, we need to imagine comparing two weeks in the same 
store that are identical in every way except temperature.
Store fixed effects remove all time‑invariant store differences.
Date fixed effects remove all national‑level shocks and seasonality.

What’s Left:
The remaining variation comes from weeks where one store happened to be 
warmer or colder than the average for that week across all stores.
That variation is plausibly more “as if random” after 
we condition on the controls.

But what about the other controls (Fuel_Price, CPI, Unemployment)?
These variables could be:

Confounders: Things that affect both temperature and 
sales (e.g., economic conditions might be correlated with weather patterns).

Mediators: Things that temperature affects, which then affect sales 
(e.g., higher temperatures could increase fuel demand and raise fuel prices).

If they are mediators, controlling for them would bias the temperature 
coefficient (because we would be blocking part of the causal path).
If they are confounders, we should control for them.

In practice, it’s hard to know for sure.
I will include them in the main model because they capture some local 
economic variation that is not already absorbed by 
date fixed effects (date FE only removes national trends).
Later, as a robustness check, I will run the model without them to see 
if the temperature estimate changes.

One detail is important here, because the model includes Date fixed effects, 
any purely national movement in fuel prices, CPI, or unemployment is already 
absorbed by the week dummies. These controls only matter through differences 
across stores within the same week, which helps account for local economic 
conditions that could be correlated with local weather shocks.


Causal story (Informal DAG):

Temperature → May affect sales directly 
(people might shop less when very hot or cold).

Temperature → May affect fuel prices 
(demand for energy) which could affect sales.

Temperature → May affect unemployment (e.g., seasonal jobs) 
which could affect sales.

Seasonality → Affects both temperature and sales. (Date FE removes this.)

Local economic conditions → Could affect sales and might be correlated 
with temperature e.g., a heat wave might coincide with a local economic boom. 
Fuel, CPI, Unemployment help control for this.

So the main model will include Fuel_Price, CPI, Unemployment, 
plus store and date fixed effects.

```{r}

# 8. Main Fixed Effects Model

m_main <- feols(
  Log_Sales ~ Temperature + Fuel_Price + CPI + Unemployment | Store + Date,
  data = walmart,
  cluster = ~ Store + Date
)

```


```{r}

# 9. Checking for Non‑linearity

walmart <- walmart %>%
  mutate(Temp_sq = Temperature^2)

m_quad <- feols(
  Log_Sales ~ Temperature + Temp_sq + Fuel_Price + CPI + 
    Unemployment | Store + Date, data = walmart,
  cluster = ~ Store + Date
)

```

To check whether the temperature effect is nonlinear, I added a 
quadratic term (Temperature²). In my results, the quadratic term is 
statistically significant, which suggests the relationship is not 
perfectly linear. That said, the quadratic coefficient is extremely small, 
so the curvature is mild in practice. We can say that sales change 
with temperature, but the shape of that relationship does not bend sharply 
across the temperature range in this dataset. 
So I treat the quadratic term as evidence of a small nonlinear pattern, 
but the overall temperature effect is still modest.

The quadratic coefficient appears as 0.000 in the table due to rounding however, 
it is small in magnitude but statistically significant.

```{r}

# 10. Check for Lagged Effects

walmart <- walmart %>%
  group_by(Store) %>%
  arrange(Date) %>%
  mutate(
    Temp_lag1 = lag(Temperature, 1)
  ) %>%
  ungroup()

m_lag <- feols(
  Log_Sales ~ Temperature + Temp_lag1 + Fuel_Price + CPI + 
  Unemployment | Store + Date, data = walmart,
  cluster = ~ Store + Date
)

summary(m_lag)

```
I also tested whether temperature affects sales with a delay by adding 
last week’s temperature to the model. The lagged temperature coefficient is 
statistically significant at the 1% level. This suggests that temperature may 
have a small carryover effect into the following week. 
The magnitude of the coefficient is about 0.00157 which implies that a 
one-degree increase in temperature last week is associated with roughly a 0.157% 
increase in sales this week, holding store and week fixed effects constant.

Even though the effect is statistically significant, it remains economically 
small. Overall, the results suggest that temperature influences sales mainly in 
the current week, with a modest additional impact from the previous week.

```{r}

# 11. Robustness Checks
# 11.1 Without economic controls
# If fuel price, CPI, and unemployment are mediators, we should see the 
# temperature coefficient change when we drop them.

m_no_controls <- feols(
  Log_Sales ~ Temperature | Store + Date,
  data = walmart,
  cluster = ~ Store + Date
)

```


```{r}

# 11.2 Adding store‑specific time trends
# Sometimes stores have their own trends 
# (e.g., a store growing faster than others).
# I add store‑specific linear time trends to see if that changes the result.

walmart <- walmart %>%
  group_by(Store) %>%
  mutate(week_id = as.numeric(Date - min(Date)) / 7) %>%   
  ungroup()

m_trend <- feols(
  Log_Sales ~ Temperature + Fuel_Price + CPI + Unemployment | Store[week_id] + 
  Date, data = walmart, cluster = ~ Store + Date
)

```


```{r}

# 12. Results Tables

models_1 <- list(
  "Main" = m_main,
  "No controls" = m_no_controls,
  "Quadratic" = m_quad
)

models_2 <- list(
  "With lag" = m_lag,
  "Store trends" = m_trend
)

common_args <- list(
  stars = c("*" = .1, "**" = .05, "***" = .01),
  coef_map = c(
    "Temperature" = "Temperature",
    "Temp_sq" = "Temperature²",
    "Temp_lag1" = "Temperature (1-week lag)"
  ),
  gof_map = c("nobs", "r.squared", "vcov.type"),
  notes = paste(
    "All models include Store and Date fixed effects, except where noted.",
    "Standard errors clustered by Store and Date."
  )
)

do.call(modelsummary, c(list(
  models_1,
  title = "Regression Results (1/2): Main, No Controls, Quadratic"
), common_args))

do.call(modelsummary, c(list(
  models_2,
  title = "Regression Results (2/2): With Lag and Store Trends"
), common_args))


```


```{r}

# 13. Interpreting the Results

b_temp <- coef(m_main)["Temperature"]
pct_change <- 100 * (exp(b_temp) - 1)

ci <- confint(m_main, parm = "Temperature")
ci_pct <- 100 * (exp(ci) - 1)

cat("Temperature coefficient (log points): ", round(b_temp, 6), "\n", sep = "")
cat("Approx. percent change in sales per +1°F: ", 
    round(pct_change, 3), "%\n", sep = "")
cat("95% CI in percent terms: [", round(ci_pct[[1]], 3), "%, ", 
    round(ci_pct[[2]], 3), "%]\n", sep = "")

```

In the log model, the temperature coefficient translates into about 
a 0.199% increase in weekly sales for a 1°F increase in temperature, 
holding the other controls and fixed effects constant. 
The 95% confidence interval is roughly 0.076% to 0.323%, so the estimate is 
positive and fairly precise statistically. 
Even so, the effect size is small in business terms: a one-degree change in 
temperature is not expected to meaningfully change sales on its own.

Now we look at the other models.

No controls gives a slightly larger coefficient (about 0.0020). 
This suggests that adding Fuel_Price, CPI, and Unemployment explains a 
small part of the variation, but the temperature effect remains very similar.

Quadratic: The squared term is extremely small, and it does not change 
the main conclusion. Any non linearity appears mild in practical terms.

With lag: The lagged temperature coefficient is statistically significant, 
which suggests a small carryover effect into the next week. 
However, the magnitude is still modest, so the economic impact remains small.

Store trends: The coefficient barely changes (around 0.0016), which suggests
store-specific trends are not driving the result.

Overall, the temperature estimate is stable across specifications, and the 
main conclusion stays the same: the effect is positive but small.

```{r}

# 14. Visual Check of the Residual Relationship
# This plot shows the relationship that the fixed‑effects model actually uses, 
# after removing store effects, date effects, and controls from both variables.

resid_sales <- resid(feols(Log_Sales ~ Fuel_Price + CPI + 
                             Unemployment | Store + Date, data = walmart))
resid_temp  <- resid(feols(Temperature ~ Fuel_Price + CPI + 
                             Unemployment | Store + Date, data = walmart))

plot_df <- tibble(resid_sales = resid_sales, resid_temp = resid_temp)

ggplot(plot_df, aes(x = resid_temp, y = resid_sales)) +
  geom_point(alpha = 0.10, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  labs(
    title = "Residualized Relationship (Matches the FE Model)",
    subtitle = "After removing Store FE, Date FE, and economic controls",
    x = "Temperature residual (°F)",
    y = "Log sales residual"
  ) +
  theme_minimal()

```

The line is slightly upward but very shallow, consistent with 
a small positive effect.

15. Answer to the Research Question
Do higher temperatures causally affect Walmart weekly sales?

Based on the fixed‑effects models:

The estimated effect is positive but very small, about a 0.20% increase 
in sales per 1°F (roughly 0.199% in the main model).

This result is robust to dropping controls, adding a quadratic term, 
including a lag and adding store‑specific trends.

The effect is economically small compared to normal seasonal 
swings or holiday weeks.

So while temperature does have a statistically detectable effect, it is not 
large enough to be a major concern for Walmart’s overall sales forecasting.

16. Limitations
Even with store and date fixed effects, we cannot completely rule out 
local,time‑varying confounders. For example, a local festival or a severe 
storm could affect both temperature perceptions and shopping behavior, and 
those events might not be fully captured by date fixed effects, 
because they hit only one store.
That kind of bias would be small and unlikely to overturn the conclusion, 
but it’s worth mentioning.

Also, the data covers only 45 stores over about 2.5 years, results might 
differ in other regions or longer periods.

17. Executive Summary
Sales and temperature both follow seasonal patterns, so a simple comparison 
is misleading. After using store and week fixed effects to control for those 
patterns, I find that a 1°F warmer week is associated with only about 
a 0.20% increase in sales (about 0.199% in the main model).
That’s a tiny effect, much smaller than the usual holiday bumps.
Walmart probably does not need to worry about gradual temperature 
changes when planning weekly sales.
